Software Architectures for Integrating Workflow Engines in Science Gateways

### 2016-05-13 paper.tex
Update paper.tex

\begin{abstract}
  We investigate six software architectures commonly used to integrate
  workflow engines in science gateways. In \emph{tight integration},
  the workflow engine shares software components with the science
  gateway. In \emph{service invocation}, the engine is isolated and
  invoked through a specific software interface. In \emph{task
    encapsulation}, the engine is wrapped as a computing task executed
  on the infrastructure. In the \emph{pool model}, the engine is
  bundled in an agent that connects to a central pool to fetch and
  execute workflows. In \emph{nested workflows}, the engine is
  integrated as a child process of another engine. In \emph{workflow
    import}, the engine is integrated through workflow language
  conversion. We describe and evaluate these architectures with
  metrics that measure integration complexity, robustness,
  extensibility, scalability and support for meta-workflows and
  fine-grained debugging\closednote{Naj}{Please be precise, what are
    they?}{Tristan}{Done, thanks}. Results provide insights for
  science gateway architects and workflow engine developers. Tight
  integration and task encapsulation are the easiest to integrate and
  are most robust. Extensibility is equivalent in most
  architectures. The pool model is the most scalable one and
  meta-workflows are only available in nested workflows and workflow
  import. \closednote{Naj}{I find this abstract to be vague, it can be made
    more concise and poignant}{Tristan}{Added a description of the
    architectures, following @pbellec's suggestion.}
\end{abstract}

@@

Workflow engines are critical for the efficient and transparent
exploitation of distributed infrastructures in the ecosystem of tools
and services offered by science gateways. Several software
architectures can be adopted to integrate workflow engines in science
gateways, with important consequences on the development effort
required and resulting system.

This paper describes, provides examples and compares such
architectures, based on system-independent representations of their
main components and interactions. It is informed by our experince in
the development and sustained operation of the CBRAIN~\cite{SHER-14}
and VIP~\cite{GLAT-13} science gateways for medical image analysis
during the past 7 years; as well as by lessons learned from several
science gateway and workflow projects such as
SHIWA\footnote{\url{http://www.shiwa-workflow.eu}}.  In this section
we provide background information and definitions of 
workflow engines, science gateways and infrastructures. In
Section~\ref{sec:architectures}, we describe six architectures within
a consistent framework that underlines the functional interactions
between their main software components. In
Section~\ref{sec:evaluation}, the different architectures are
evaluated with metrics measuring integration complexity, robustness,
extensibility, scalability and other specific features. We finally
compare the architectures and highlight recommendations for science
gateway architects and workflow engine developers.

@@

At the same time, toolboxes have been emerging in various scientific
domains to facilitate the assembly of software components in
consistent ``pipelines''. In neuroimaging, our primary domain of
interest, tools such as Nipype (Neuroimaging in Python, Pipelines and
Interfaces~\cite{gorgolewski2011nipype}), PSOM (Pipeline System for
Octave and Matlab~\cite{bellec2012pipeline}), RPPL~\cite{1174106} and
PMP (Poor Man's Pipeline~\cite{Ad-DabbaghY2006}), SPM (Statistical
Parametric Mapping~\cite{ashburner2011spm}) and FSL (FMRIB Software
Library~\cite{Jenkinson2012782}) provide abstractions and functions to
handle the data and computing flow between processes implemented in a
variety of programming languages. Such tools were interfaced to
computing infrastructures, in particular
clusters\closednote{Tristan}{Reverted from Marc's 'cluster
  scheduler'}{}{}, to execute tasks at a high throughput. Some of
these tools also support advanced features such as provenance
tracking, \closednote{Naj}{should we not mention PMP which is used for
  CIVET?}{Tristan}{Done, as discussed.}  redundancy detection across
analyses to avoid re-computation, and so on. A wide-array of
workflows have been implemented using these pipelines and are now
shared across neuroimaging groups world-wide. These domain-specific
engines represent a tremendous opportunity for science gateways to
leverage existing tools and applications. They nicely complement
e-Science engines that are more oriented towards the exploitation of
distributed computing infrastructures, in particular grids and clouds.

In this paper, a
\emph{workflow engine} (also abbreviated \emph{engine}) is a piece of
software that submits interdependent computing \emph{tasks} to an
\emph{infrastructure} (local server, cluster, grid or cloud) based on
a workflow description (a.k.a \emph{workflow}) and using input data
that may consist of files, database entries or simple parameter
values. Although simplistic, this definition covers both e-Science workflow systems and
domain-specific pipelines.\closednote{Marc-e}{To remain consistent with above descriptions}{Tristan}{yes!} Some workflow engines, usually e-Science
ones, may transfer data across the infrastructure, and others, usually
domain-specific ones, may leave this role to external processes. Workflows
may be expressed in any language, including high-level XML or JSON
dialects such as Scufl or Hyperflow, and low-level scripting languages
such as Bash.

@@

Various science gateways have been developed, including complete
frameworks such as Apache Airavata~\cite{marru2011apache}, the Catania
Science Gateway Framework~\cite{Ardizzone2012} and
WS-PGRADE/gUSE~\cite{Kacsuk2012}. Numerous science gateways were built
using such frameworks~\cite{kacsuk2014science,ardizzone2012decide} or
as standalone systems~\cite{SHER-14,GLAT-13}. Most of these frameworks
and instances include one or several workflow engines.

@@

Integration between workflow engines and science gateways varies
across systems. Some science gateways are tailored to a particular
engine while other ones are more general and host applications
executed by different types of engines. Some gateways may not even use any
workflow engine and instead focus on the execution of simple tasks.

Extensibility is an important property of the integration. New
workflows are added frequently, different types and versions of
workflow engines may be integrated over time, and different kinds of
infrastructure can be targeted.

Scalability is also a crucial concern for such multi-user,
high-throughput systems. To this end, science gateways may balance the
load among different instances of the same engine, and use advanced
task scheduling policies on the infrastructure to improve performance,
fault-tolerance and fairness among users.

Robustness is highly desirable as well since it is key to a
good user experience. Simple architectures facilitate the
implementation effort towards robust interactions which, in turn, have
a positive impact on characteristics such as gateway predictability,
transparency, reliability, traceability and
reproducibility. \closednote{Marc-Etienne}{I find that something is
  weak in this last paragraph. Robustness will help with transparency,
  sure, but it seems only a small facet.  Also, the text seem to
  timidly favor architectures with less components/interactions
  without much explanation... Complex systems can be robust, but at
  greater costs...  Here is one: "Architectures showing reduced system
  complexity will greatly facilitate the implementation effort towards
  robust interactions which, in turn, have a positive impact on
  characteristics such as gateway predictability, transparence,
  general reliability and results traceability and reproducibility;
  characteristics that are also key to a good user
  experience."}{Tristan}{Agreed, fixed as
  suggested.}\closednote{Tristan}{Work on the this sentence again.}{Tristan}{done.}

Specific features may also be available in science gateways, such as
data vizualisation and quality control, workflow edition, debugging
instruments, social tools among users, etc.

@@

\paragraph{Scalability} The service architecture supports multiple
engine instances through multiple workflow services. In VIP for
instance, this feature has been available from release 1.17 (April
2016). A basic load-balancing mechanism is available that sends new
workflow executions to the engine instance that has the least active
executions. To avoid ``black-hole'' syndromes created by failing
engine instances, engine instances are automatically disabled when
workflows cannot be submitted to them. Adding a new engine instance,
however, requires manual intervention to declare the new instance in
the science gateway (\texttt{S$_1$}=1). Consequently, elastic engines
are difficult to implement because they require a mechanism to update
the science gateway configuration when a new engine instance is
available. Distributing the execution of a single
workflow in multiple engines is usually not possible unless the
workflow engine has specific abilities (\texttt{S$_2$}=1). The
scheduling of tasks on the infrastructure is as complex as in any
other architecture since the workflow engine might implement any kind
of scheduling policy (\texttt{S$_3$}=0).

@@

\paragraph{Scalability}
New engine instances are spawned and executed on the infrastructure as
any other task upon user submission (\texttt{S$_1$}=0). This is a
major interest of task encapsulation. Distributed engines are
not supported by default (\texttt{S$_2$}=1). Task scheduling is
slightly more complex than in the other approaches due to the special
role of the task that executes the workflow engine
(\texttt{S$_3$}=1). Indeed, the reliability of this task is critical
since all the sub-tasks in the workflow depend on it and, depending on
the recovery capabilities of the workflow engine, may need to be
resubmitted if the workflow task fails. The workflow task is also
longer than all its sub-tasks, which increases its chances of
failure. In addition, task parameters, for instance estimated
walltime, are more difficult to estimate for the workflow task than
for the sub-tasks, which may generate issues such as selection of wrong
batch queues on clusters. Finally, the interdependencies between the
workflow task and its sub-tasks may create deadlocks when there is
contention. For instance, if only 1 computing resource is
available for the science gateway and if the workflow task is
running on it and submits sub-tasks, then the sub-tasks could only
execute when the resource is available, which will never happen
because the workflow task will not complete until the sub-tasks
complete. This configuration can be generalized to an infrastructure
with $n$ resources where $n$ workflows are submitted. In practice,
however, the number of submitted workflows usually remains lower than
the number of computing resources available on this infrastructure,
which makes such deadlocks unlikely to happen.

@@

\paragraph{Scalability} New engine instances only require new agents,
which is easily automated (\texttt{S$_1$=0}) and by design very
suitable for elastic computing. For instance, auto-scaling rules can
be implemented to start new agents when the workload in the science
gateway exceeds a certain threshold~\cite{lorido2012auto}. Distributed
engines are not available by default (\texttt{S$_2$=1}) and task
scheduling is as complex as in any other architecture
(\texttt{S$_3$=0}).

@@

Tight integration and task encapsulation are the simplest
architectures to integrate, followed by service integration, workflow import (with
service invocation) and pool. Nested workflows (with service
invocation) requires more integration than the other
architectures. Robustness roughly leads to the same ordering of
architectures, with tight integration and task encapsulation in the top
group, service integration and workflow import (with service
invocation) close behind, pool in a third group, and nested workflows
(with service invocation) at the end. This ordering is consistent
across metrics; it reflects the global complexity of the
architectures.

Regarding extensibility, most architectures are overall comparable,
except nested workflows (with service invocation) which is
significantly behind. This is explained by the complexity of the
nested workflow architecture, with 2 infrastructures and 2 workflow
services. Task encapsulation and workflow import (with service invocation)
perform a bit better when integrating new engines (\texttt{E$_1$}),
which is interesting given the current profusion of engines with
different characteristics.  However, task encapsulation, workflow import
(with service invocation) and tight integration perform worse than the
others when adding engine versions (\texttt{E$_2$}) due to the need to
update infrastructure (task encapsulation), science gateway (tight
integration), or workflow converter (workflow import). Workflow import
performs worse than the others when integrating new workflows
(\texttt{E$_3$}) because of the language conversion step. All
architectures except nested workflows perform the same when
integrating new infrastructures (\texttt{E$_4$}).

The pool architecture is overall the most scalable, which is not a
surprise since it was designed precisely for scalability. Tight
integration is the least scalable and all the other architectures
perform the same overall. The global scalability score, however, should not
mask the unique characteristics of architectures regarding this
criterion. Nested workflows are the only architecture that can easily
accommodate distributed workflow execution, which can be critical in
some cases. At the same time, the scheduling constraints created by
task encapsulation and nested workflows could well become
showstoppers depending on the type of targeted
infrastructure. Non-reliable infrastructures, for example, could
hardly cope with workflow engines being wrapped in computing tasks as
done in task encapsulation. The availability of multiple engine instances
could also become a critical feature for science gateways with
important workloads, which would favor task encapsulation and pool.

Differences in other specific features should not be neglected. Nested
workflows (with service invocation) and workflow import (with service
invocation) are the only architectures that support meta-workflows,
which may be interesting in some cases. Also, task encapsulation and nested
workflows may complexify fine-grained debugging. \closednote{Rafael}{I would
  suggest to provide some more examples when this could be
  interesting.}{Tristan}{Done.}  The availability of fine-grained debugging
information may also be critical to efficient user support.

@@

A few limitations remain that should be considered when using the
results of our evaluation. First of all, the evaluation is done at a
quite high-level of abstraction that may be a bit artificial when
applied to real systems. Abstracting architectures from their
implementation in specific systems is useful to compare them, but we
also realize that it comes at the cost of realism. In particular, the
presented architectures are theoretical patterns that may be blended
in actual systems. The distinction between tight integration and
service invocation, for instance, may not always be that clear in
practice. Service invocation may also be combined with task
encapsulation in some cases. However, the criteria discussed in this
work would still apply to broadly compare and categorize such cases of
hydrid architectures.

Moreover, all the interactions involved in the architectures were
treated equally while some of them are obviously more complex
than others. For example, interaction \texttt{g} (workflow language
conversion) is clearly more complex than interaction \texttt{d}
(service invocation), and may create much more reliability issues as
well. However, quantifying the robustness and amount of development
associated with each interaction and software component cannot
realistically be done without entering into the details of a
particular system.

The particular case of interaction \texttt{g} (workflow language
conversion) should be further discussed since it is very likely that
this interaction could not be easily generalized to any workflow
language. For instance, converting FSL, PSOM or Nipype pipelines to
any other workflow language is problematic because these engines rely
on general-purpose programming languages such as Bash, Octave and
Python, which are much richer than scientific workflow languages.

The abstract nested workflows and workflow import patterns were
instantiated with service invocation so that they can be analyzed in
the same framework as the other architectures. Other types of
instantiation, for instance nested workflows with task encapsulation,
could also be envisaged. We chose to limit ourselves to instantiations
with service invocation because the resulting architectures are
implemented in real systems, and because service invocation is largely
used. Nevertheless, it could be interesting to explore other types of
instantiations, in particular nested workflows with task
encapsulation.

The particular technical or historical context of a science gateway
project may obviously influence the choice of an architecture to
integrate workflow engines. For instance, some workflow engines may be
already available as web services, which would tend to favor service
invocation, and other science gateways may have strongly tested task
and data control features (interactions \texttt{b} and \texttt{c}),
which would favor task encapsulation. Similarly, adding a new type of engine
may facilitate integration of a new infrastructure when interactions
\texttt{b} and \texttt{c$_2$} are already available. The migration
cost between architectures has been ignored as well.

@@

% Summary of contributions
We reviewed architectures used to integrate workflow engines in
science gateways, we described them in a system-independent framework
suitable for comparison, we illustrated them on real systems, we
developed quantitative metrics for evaluation, and we discussed the
pros and cons of all the presented architectures based on these metrics.
\closedanswerednote{Rafael}{One thing that is missing in the
  conclusion or the limitations section is to state that this work
  does not intend to quantify the amount of work required to develop
  science gateways using a specific architecture, but the goal is to
  highlight what are the main components and interactions for each
  architecture, and the main advantages and drawbacks. This should
  also be stated in the introduction.}{Tristan}{This is in
  contradiction with the metric on 'integration effort'. What do you
  think?}{Tristan}{Fixed, as discussed in github \#8.}

% Impact of the work
To the best of our knowledge, our work is the first of its kind. So
far, the literature on science gateways and workflow engines has
focused on the description of particular systems, or on the
presentation of a particular architecture. For instance, the work
in~\cite{Gannon2005} discusses service-oriented architectures for
science gateways with workflow engines.  Instead, our analysis
abstracts and evaluates architectural patterns independently from any
particular system. It provides general insight to integrate science
gateways and workflow engines, which we hope will be useful for
software architects.

>>>>

\begin{abstract}
  We investigate six software architectures commonly used to integrate
  workflow engines in science gateways. (1) In \emph{tight
    integration}, the workflow engine shares software components with
  the science gateway; (2) In \emph{service invocation}, the engine is
  isolated and invoked through a specific software interface; (3) In
  \emph{task encapsulation}, the engine is wrapped as a computing task
  executed on the infrastructure; (4) In the \emph{pool model}, the
  engine is bundled in an agent that connects to a central pool to
  fetch and execute workflows; (5) In \emph{nested workflows}, the
  engine is integrated as a child process of another engine; (6) In
  \emph{workflow import}, the engine is integrated through workflow
  language conversion. We describe and evaluate these architectures
  with metrics that measure integration complexity, robustness,
  extensibility, scalability and support for meta-workflows and
  fine-grained debugging\closednote{Naj}{Please be precise, what are
    they?}{Tristan}{Done, thanks}. Results provide insights for
  science gateway architects and workflow engine developers. Tight
  integration and task encapsulation are the easiest to integrate and
  are most robust. Extensibility is equivalent in most
  architectures. The pool model is the most scalable one and
  meta-workflows are only available in nested workflows and workflow
  import. \note{Naj}{I find this abstract to be vague, it can be made
    more concise and poignant}
\end{abstract}

@@

Workflow engines are critical for the efficient and transparent
exploitation of distributed infrastructures in the ecosystem of tools
and services offered by science gateways. Several software
architectures can be adopted to integrate workflow engines in science
gateways, however development effort required and the resulting system
vary across different architectures.

This paper describes, provides examples and compares such
architectures, based on system-independent representations of their
main components and interactions. It is informed by our experince in
the development and sustained operation of the CBRAIN~\cite{SHER-14}
and VIP~\cite{GLAT-13} science gateways for medical image analysis
during the past 7 years; as well as by lessons learned from several
science gateway and workflow projects such as
SHIWA\footnote{\url{http://www.shiwa-workflow.eu}}.  In this section
we provide background information and definitions of different
workflow engines, science gateways and infrastructures. In
Section~\ref{sec:architectures}, we describe six architectures within
a consistent framework that underlines the functional interactions
between their main software components. In
Section~\ref{sec:evaluation}, the different architectures are
evaluated with metrics measuring integration complexity, robustness,
extensibility, scalability and other specific features. We finally
compare the architectures and highlight recommendations for science
gateway architects and workflow engine developers.

@@

At the same time, toolboxes have been emerging in various scientific
domains to facilitate the assembly of software components in
consistent ``pipelines''. In neuroimaging, our primary domain of
interest, tools such as Nipype (Neuroimaging in Python, Pipelines and
Interfaces~\cite{gorgolewski2011nipype}), PSOM (Pipeline System for
Octave and Matlab~\cite{bellec2012pipeline}), RPPL~\cite{1174106} and PMP
(Poor Man's Pipeline~\cite{Ad-DabbaghY2006}), SPM (Statistical Parametric
Mapping~\cite{ashburner2011spm}) and FSL (FMRIB Software
Library~\cite{Jenkinson2012782}) provide abstractions and functions to
handle the data and computing flow between processes implemented in a
variety of programming languages. Such tools were interfaced to
computing infrastructures, in particular
clusters\closednote{Tristan}{Reverted from Marc's 'cluster
  scheduler'}{}{}, to execute tasks at a high throughput. Some of
these tools also support advanced features such as provenance tracking
(Nipype, PSOM), \closednote{Naj}{should we not mention PMP which is
  used for CIVET?}{Tristan}{Done, as discussed.}
redundancy detection across analyses to avoid re-computation (PSOM),
and so on \note{naj}{please remove so on, and be more precise?}. A wide-array \note{Naj}{no need for hyphenation here} of workflows have been implemented using these
pipelines and are now shared across neuroimaging groups
world-wide. These domain-specific engines represent a tremendous
opportunity for science gateways to leverage existing tools and
applications. They nicely complement e-Science engines that are more
oriented towards the exploitation of distributed computing
infrastructures, in particular grids and clouds.

In this paper, a
\emph{workflow engine} (also abbreviated \emph{engine}) is a piece of
software that submits interdependent computing \emph{tasks} to an
\emph{infrastructure} (local server, cluster, grid or cloud) based on
a workflow description (a.k.a \emph{workflow}) and\note{Naj}{I think and is redundant here, you can replace is with a comma} using input data
that may consist of files, database entries or simple parameter
values. Although simplistic, this definition covers both e-Science workflow systems and
domain-specific pipelines.\closednote{Marc-e}{To remain consistent with above descriptions}{Tristan}{yes!} Some workflow engines, usually e-Science
ones, may transfer data across the infrastructure, and others, usually
domain-specific ones, may leave this role to external processes. Workflows
may be expressed in any language, including high-level XML or JSON
dialects such as Scufl or Hyperflow, and low-level scripting languages
such as Bash.

@@

Various science gateways have been developed, including complete
frameworks such as Apache Airavata~\cite{marru2011apache}, the Catania
Science Gateway Framework~\cite{Ardizzone2012} and
WS-PGRADE/gUSE~\cite{Kacsuk2012}. Numerous science gateways were built
using such frameworks~\cite{kacsuk2014science,ardizzone2012decide} or
as standalone systems~\cite{SHER-14,GLAT-13}. Most of these science
gateway frameworks and instances include one or several workflow
engines.

@@

Integration between workflow engines and science gateways varies
across systems. Some science gateways are tailored to a particular
engine, while other ones are more general and host applications
executed by different types of engines. Some gateways may not even use any
workflow engine and focus on the execution of simple tasks.

Extensibility is an important property of the integration, as new
workflows are added frequently, different types and versions of
workflow engines may be integrated over time, and different kinds of
infrastructure can be targeted. Scalability is also a crucial concern
for such multi-user systems that manage substantial amounts of data
and execute computationally-intensive analyses. To this end, science
gateways may balance the load among different instances of the same
engine. They may also use advanced task scheduling policies on the
infrastructure, e.g. to ensure fairness among
users~\cite{FERR-14,CPE:CPE3708} or to improve
fault-tolerance~\cite{wilkins2008teragrid,FERR-13}.

Robustness is a highly desirable characteristic as well. Architectures
showing reduced system complexity will greatly facilitate the
implementation effort towards robust interactions which, in turn, will have
a positive impact on characteristics such as gateway predictability,
transparence, general reliability and results traceability and
reproducibility; characteristics that are also key to a good user
experience. \closednote{Marc-Etienne}{I find that something is weak in
  this last paragraph. Robustness will help with transparency, sure,
  but it seems only a small facet.  Also, the text seem to timidly
  favor architectures with less components/interactions without much
  explanation... Complex systems can be robust, but at greater
  costs...  Here is one: "Architectures showing reduced system
  complexity will greatly facilitate the implementation effort towards
  robust interactions which, in turn, have a positive impact on
  characteristics such as gateway predictability, transparence,
  general reliability and results traceability and reproducibility;
  characteristics that are also key to a good user
  experience."}{Tristan}{Agreed, fixed as
  suggested.}\note{Tristan}{Work on the this sentence again.}

Specific features may also be available in science gateways, such as
data vizualisation and quality control, workflow edition, debugging
instruments, social tools among users, etc. \note{Naj}{I think using etc after such as, for example and e.g. is not correct. --sorry for being pedantic}

@@

\paragraph{Scalability} The service architecture supports multiple
engine instances through multiple workflow services. In VIP for
instance, this feature has been available from release 1.17 (April
2016). A basic load-balancing mechanism is available that sends new
workflow executions to the engine instance which has the least active
executions. To avoid ``black-hole'' syndromes created by failing
engine instances, engine instances are automatically disabled when
workflows cannot be submitted to them. Adding a new engine instance,
however, requires manual intervention to declare the new instance in
the science gateway (\texttt{S$_1$}=1). Consequently, elastic engines
are difficult to implement because they require a mechanism to update
the science gateway configuration when a new engine instance is
available. Distributing the execution of a single
workflow in multiple engines is usually not possible unless the
workflow engine has specific abilities (\texttt{S$_2$}=1). The
scheduling of tasks on the infrastructure is as complex as in any
other architecture since the workflow engine might implement any kind
of scheduling policy (\texttt{S$_3$}=0).

@@

\paragraph{Scalability}
New engine instances are spawned and executed on the infrastructure as
any other task upon user submission (\texttt{S$_1$}=0). This is a
major interest of task encapsulation. Distributed engines are
not supported by default (\texttt{S$_2$}=1). Task scheduling is
slightly more complex than in the other approaches due to the special
role of the task that executes the workflow engine
(\texttt{S$_3$}=1). Indeed, the reliability of this task is critical
since all the sub-tasks in the workflow depend on it and, depending on
the recovery capabilities of the workflow engine, may need to be
resubmitted if the workflow task fails. The workflow task is also
longer than all its sub-tasks, which increases its chances of
failure. In addition, task parameters, for instance estimated
walltime, are more difficult to estimate for the workflow task than
for the sub-tasks, which may generate issues such as selection of wrong
batch queues on clusters \note{Naj}{The lay reader wonders why? an example like the one offered for interdependencies can help}. Finally, the interdependencies between the
workflow task and its sub-tasks may create deadlocks when there is
contention. For instance, if only 1 computing resource is
available for the science gateway and if the workflow task is
running on it and submits sub-tasks, then the sub-tasks could only
execute when the resource is available, which will never happen
because the workflow task will not complete until the sub-tasks
complete. This configuration can be generalized to an infrastructure
with $n$ resources where $n$ workflows are submitted. In practice,
however, the number of submitted workflows usually remains lower than
the number of computing resources available on this infrastructure,
which makes such deadlocks unlikely to happen. \note{Naj}{so is this the solution to fixing the above mentioned limitations with scalability?}

@@

\paragraph{Scalability} New engine instances only require new agents,
which is easily automated (\texttt{S$_1$=0}) and by design very
suitable for elastic computing.\note{Naj}{I feel the concept of elastic computing should be mentioned in introductory sections} For instance, auto-scaling rules can
be implemented to start new agents when the workload in the science
gateway exceeds a certain threshold~\cite{lorido2012auto}. Distributed
engines are not available by default (\texttt{S$_2$=1}) and task
scheduling is as complex as in any other architecture
(\texttt{S$_3$=0}).

@@

Tight integration and task encapsulation are the simplest
architectures to integrate, followed by service integration, workflow import (with
service invocation) and pool. Nested workflows (with service
invocation) require more integration than the other
architectures. Robustness roughly leads to the same ordering of
architectures, with tight integration and task encapsulation in the top
group, service integration and workflow import (with service
invocation) close behind, pool in a third group, and nested workflows
(with service invocation) at the end. This ordering is consistent
across metrics; it reflects the global complexity of the
architectures.

Regarding extensibility, most architectures are overall comparable,
except nested workflows (with service invocation) which are
significantly behind. This is explained by the complexity of the
nested workflow architecture, with 2 infrastructures and 2 workflow
services. Task encapsulation and workflow import (with service invocation)
perform slightly better when integrating new engines (\texttt{E$_1$}),
which is interesting given the current profusion of engines with
different characteristics.\note {Naj}{what is interesting?}  However, task encapsulation, workflow import
(with service invocation) and tight integration perform worse than the
others when adding engine versions (\texttt{E$_2$}), due to the need to
update infrastructure (task encapsulation), science gateway (tight
integration), or workflow converter (workflow import). Workflow import
performs worse than the others when integrating new workflows
(\texttt{E$_3$}) because of the language conversion step. All
architectures except nested workflows perform the same when
integrating new infrastructures (\texttt{E$_4$}).

The pool architecture is overall the most scalable, which is not surprising since it is designed precisely for scalability. Tight
integration is the least scalable and all the other architectures
perform the same overall. The global scalability score, however, should not
mask \note{Naj}{what do you mean by mask?}the unique characteristics of architectures regarding this
criterion. Nested workflows are the only architecture that can easily
accommodate distributed workflow execution, which can be critical in
some cases. At the same time, the scheduling constraints created by
task encapsulation and nested workflows could well become
showstoppers \note{Naj}{please don't use these from-film-marketing words?! :-) }depending on the type of targeted
infrastructure. Non-reliable infrastructures \note{Naj}{you mean by the showstoppers, i.e. the nested workflows which can accommodate distributed executions but are limited by scheduling constraints?} , for example, could
hardly cope with workflow engines being wrapped in computing tasks as
done in task encapsulation. The availability of multiple engine instances
could also become a critical feature for science gateways with
important workloads, which would favor task encapsulation and pool. \note{Naj}{I feel this paragraph is a bit vague, and doesn't distill and contextualize the thorough work described before.}

Differences in other specific features should not be neglected. Nested
workflows (with service invocation) and workflow import (with service
invocation) are the only architectures that support meta-workflows,
which may be interesting in some cases \note{Naj}{Which?}. Also, task encapsulation and nested
workflows may complexify fine-grained debugging. \closednote{Rafael}{I would
  suggest to provide some more examples when this could be
  interesting.}{Tristan}{Done.}  The availability of fine-grained debugging
information may also be critical to efficient user support.

@@

A few limitations remain that should be considered when using the
results of our evaluation. First, our evaluation methods aimed to provide comparative metrics, however the high-level of abstraction used to derive these metrics hinders the complexity of real systems to some extent.
applied to real systems.\note{Naj}{Here, you were bashing your article harshly, so I rewrote it}. In particular, the
presented architectures are theoretical patterns that may be blended
in actual systems.\note{Naj}{something about "theoretical pattern", and "blending" is jarring, aare these common words in the computer sci context?} The distinction between tight integration and
service invocation, for instance, may not always be that clear in
practice. Service invocation may also be combined with task
encapsulation in some cases. However, the criteria discussed in this
work would still apply to broadly compare and categorize such cases of
hydrid architectures.

Moreover, all the interactions involved in the architectures were
treated equally while some of them are obviously more complex
than others. For example, interaction \texttt{g} (workflow language
conversion) is clearly more complex than interaction \texttt{d}
(service invocation), and may create much more reliability issues as
well. However, without entering into the details of a
particular system,  quantification of the robustness or the amount of development
associated with each interaction and software component will not be precise.

The particular case of interaction \texttt{g} (workflow language
conversion) should be further discussed \note{Naj}{so are you discussing it here, or in the future?}since 
this interaction may not be easily generalized to any workflow
language. For instance, converting FSL, PSOM or Nipype pipelines to
any other workflow language is problematic because these engines rely
on general-purpose programming languages such as Bash, Octave and
Python, which are much richer than scientific workflow languages.

The abstract nested workflows and workflow import patterns were
instantiated with service invocation so that they can be analyzed in
the same framework as the other architectures. Other types of
instantiation, for instance nested workflows with task encapsulation,
could also be envisaged. We chose to limit ourselves to instantiations
with service invocation because the resulting architectures are
implemented in real systems, and because service invocation is largely
used. Nevertheless, it could be interesting to explore other types of
instantiations, in particular nested workflows with task
encapsulation.\note{Naj}{I think in the discussion, you may want to close the circle on a statement in introduction, which referred to your 7-year experience with CBRAIN and VIP}

The particular technical or historical context of a science gateway
project may obviously influence the choice of an architecture to
integrate workflow engines. For instance, some workflow engines may be
already available as web services\note{Naj}{This is a good place to offer some examples}, which would tend to favor service
invocation, and other science gateways may have strongly tested task
and data control features (interactions \texttt{b} and \texttt{c}),
which would favor task encapsulation. Similarly, adding a new type of engine
may facilitate integration of a new infrastructure when interactions
\texttt{b} and \texttt{c$_2$} are already available. The migration
cost between architectures has been ignored as well.

@@

% Summary of contributions
We have systematically reviewed architectures used to integrate workflow engines in
science gateways. These architectures are described in a system-independent framework
suitable for comparison, illustrated on real systems, and evaluated using novel quantitative metrics that allow simple comparison across architectures. We have discussed the
pros and cons of all the presented architectures based on these metrics.
\closedanswerednote{Rafael}{One thing that is missing in the
  conclusion or the limitations section is to state that this work
  does not intend to quantify the amount of work required to develop
  science gateways using a specific architecture, but the goal is to
  highlight what are the main components and interactions for each
  architecture, and the main advantages and drawbacks. This should
  also be stated in the introduction.}{Tristan}{This is in
  contradiction with the metric on 'integration effort'. What do you
  think?}{Tristan}{Fixed, as discussed in github \#8.}

% Impact of the work
To the best of our knowledge, our work is the first of its kind.\note{Naj}{instead of its kind, explain what kind it is, e.g. systematic review of all ...} So
far, the literature on science gateways and workflow engines has
focused on the description of particular systems, or on the
presentation of a particular architecture. For instance, the work
in~\cite{Gannon2005} discusses service-oriented architectures for
science gateways with workflow engines.  Instead, our analysis
abstracts and evaluates architectural patterns independently from any
particular system. It provides general insight to integrate science
gateways and workflow engines, which we hope will be useful for
software architects.
